---
title: "Check pedigrees after they've been simulated"
output: pdf_document
date: "2023-02-03"
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

* A previous script in the workflow (`simrvped.R`) simulates
150 pedigrees with at least 4 affected members and at least one HL.
    * Pedigrees are stored in the `Outputfiles` directory in 
    plain-text files `ascertained_pedi.txt` for `i`=1,...150.
* This script identifies pedigrees that also have (i) no more 
than 6 affected members, (ii) at least one HL case who carries a cRV
and (iii) at least one NHL case. 
* Specifically, we exclude pedigrees 
with more than 6 affected members, or no HL cRV carriers, 
or no NHL cases.
* We end up with 102 pedigrees that meet the criteria
for ascertainment and for inclusion in Christina's thesis.
* From the set of 102, we choose the first 55 as the pool of pedigrees
for our simulation study.
    * Note: The number 55 is to match Christina's simulation study.

## More details on the R code

* In the R code chunk below, we initialize variables to hold
the indices of pedigrees to be excluded. Specifically,
the variable `morethan6` will hold
the indices of pedigrees with more than 6 affected
members, `noHLcarr` will
hold the indices of pedigrees in which none of the HL
cases are cRV carriers and `noNHL` will hold the indices of pedigrees with no NHL cases.
* Next, we loop over the indices
of the pedigrees, read each into R, plot it and check the
selection criteria. If a pedigree fails a criterion, it is added to the appropriate vector of indices to be excluded. 
* After the loop is finished, we keep the first 55 of the
pedigrees that were not excluded.

### Aside: An inconsistency in `sim_RVped()`'s output

* Pedigrees are stored as data frames with rows for individuals
and columns of information on each individual.
* When working with pedigrees simulated by `sim_RVped()` 
I noticed that the pedigree data structure has two columns
that indicate affection status. The first is `affected`, which is 
a logical with `TRUE` for affected individuals, `FALSE` for
unaffected individuals and `NA` for unknown affection status. 
The second is `subtype`, which is `HL` for HL-affected
individuals, `NHL` for NHL-affected individuals, and `NA` for
unaffected individuals. 
* Unfortunately, `sim_RVped()` sometimes generates pedigrees with
individuals having missing a missing value of `affected` and 
a non-missing `subtype`. It seems wrong for subtype to be known
but affection status to be unknown. I deal with this after calling `sim_RVped()` by setting the subtype variable in the
returned data structure to missing.
    * Note to self: I considered fixing the inconsistency directly in `SimRVPedigree` itself, but gave up on that idea because I don't have permissions to upload 
    revisions to CRAN as Christina is the maintainer.

```{r}
set.seed(42)
library(SimRVPedigree)
# simulation study parameters
npeds <- 150 # number of simulated pedigrees
npool <- 55 # size of pool of pedigrees we want for our simulation study
# vectors to hold indices of excluded pedigrees
morethan6 <- NULL # more than 6 disease-affected members
noHLcarr <- NULL # No HL carriers of cRV in affecteds
noNHL <- NULL # No NHL among affecteds
for(i in 1:npeds){
  cat("\n----------------Pedigree",i,"-------------------\n")
  pfile <-  paste0("Outputfiles/ascertained_ped",i,".txt")
  if(!file.exists(pfile)) {
    cat("file",pfile,"does not exist!\n")
  } else { # read the pedigree into R and check it
    ped <-  read.table(pfile)
    # ped is read in as a data frame. It also needs to have class "ped"
    # to be plotted.
    class(ped) <-c("ped","data.frame")
    plot(ped,cex=.4) 
    # Now start checking the pedigree. First check for missing affection
    # status but non-missing subtype.
    if(any(is.na(ped$affected) & !is.na(ped$subtype))) {
      cat("***ped members with missing affected and non-missing subtype -- \n")
      cat("   setting such subtypes to missing and re-writing ped to file \n")
      is.na(ped$subtype) <- is.na(ped$affected)
      write.table(ped,file=pfile)
    }
    # Check for >6 affecteds. 
    pind <- ped$affected
    pind[is.na(pind)] <- FALSE # set missing affection status to FALSE
    if(sum(pind)>6) {
      cat("***More than 6 disease-affecteds***\n")
      morethan6 <- c(morethan6,i)
    }
    # The last two checks pertain to the affecteds, so subset the
    # pedigree to just the affecteds.
    aped <- ped[pind,]
    # Check for no HL carriers of a disease allele
    aped$disease.alleles = aped$DA1+aped$DA2
    if(sum(aped$subtype=="HL" & aped$disease.alleles>0)==0){
      cat("***No HL carriers ***")
      noHLcarr <- c(noHLcarr,i)
    }
    # Lastly, check for no NHLs in the pedigree
    if(!any(aped$subtype=="NHL")) {
      cat("***No NHL subtype***\n") 
      noNHL <- c(noNHL,i)
    }
  }
}
# Print out the indices of the pedigrees to be excluded
cat("more than 6 disease-affecteds:\n")
print(morethan6)
cat("disease alleles in NHL, but none in HL:\n")
print(noHLcarr)
cat("no NHL subtype among affecteds:\n")
print(noNHL)
# Keep the pedigrees not in one of the exclusion vectors. 
keep <- setdiff(1:npeds,c(morethan6,noHLcarr,noNHL))
cat("keep peds:\n")
print(keep)
```

* Now save the indices of the first 55 usable pedigrees to use as our
pool of pedigrees for the simulation study.

```{r}
pedpool <- keep[1:npool]
save(pedpool,file="pedpool.RData")
```

## Generate the R script for the cluster.

* The above R commands run on my laptop in about 40 seconds, so it's not
really necessary to use the cluster. 
* However, if you prefer to run on the cluster, you can generate
the necessary R script by cutting-and-pasting the following into your R console:

```{r}
knitr::purl(input="checkpeds.Rmd",output="checkpeds.R")
```

* This will produce an R script containing all the code chunks
in this `Rmd` file. Delete the code chunk at the beginning of the resulting R script that sets up `knitr` (i.e., the first code chunk in this `.Rmd` file). Also delete the code chunk at
the end of the R script that contains the above `purl()` command.
