---
title: "Summaries of type-1 error rate simulations"
output:
  pdf_document:
    number_sections: true
    toc: true 
    toc_depth: 3
date: "2023-05-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

* In this document we use our PC to read in the simulation results output by
`simnull.R` and prepare a summary that can be included in the paper.
* The "output" of this document is a LaTeX table of estimated type-1 error rates 
and their SEs in Section 4.1.3.
* Before getting to the summary we read in the data and check that 
the pedigrees from our pool of pedigrees are being sampled uniformly.
* **Note:** Most other RMarkdown documents in our workflow were written as
documentation for a corresponding R script that was intended to be run on the 
cluster using a SLURM script. This RMarkdown document, like `simaltSummary.Rmd`,
is intended to be run, or "knitted" on your PC, and there are no corresponding
R and SLURM scripts.

# Read in simulation results

* `simnull.R` returns files of p-value results.
* Simulations were run on the cluster as an array job of 200 jobs, each
containing 10 studies (of three pedigrees).
* Results should be saved on the Compute Canada cluster in 
the  `/project/def-jgraham/FJdata` directory. 
* To use this script, you must first copy the
results files from the cluster to your PC. 
* Several options for transferring the results files to your PC
are discussed in Appendix A.1.4 of the Lab's Workflow document. 
* I use the `rsync` method,
for which I (i) open a terminal on my Mac and set its working directory
to be the directory that contains this .Rmd file, and (ii) run the following 
command from the terminal:
```
rsync -avz jgraham@cedar.computecanada.ca:project/FJdata/ FJdata
```
*  Batches of p-values to estimate the type-1 error rate are in the files `FJdata/pvalnullresi.csv` for i=1,...,200.
* The format of the output files was described in `simnull.Rmd` and 
this description is repeated in the Appendix of this document.
* We read the results into R in the following code chunk:

```{r}
njobs <- 200
pvalres <- read.csv("FJdata/pvalnullres1.csv") # start with first batch
for(i in 2:njobs){
  pvalres <- rbind(pvalres,read.csv(paste0("FJdata/pvalnullres",i,".csv")))
}
```


# Summary of simulations: Sampling of pedigrees

* Check that pedigrees from our pool of 55 are being sampled uniformly 
over the 2000 simulated studies.
* We take the information about which pedigrees were sampled
fromthe study information in `pvalres`.

```{r}
pedpool <- scan("FJdata/pedpool/pedpool.txt") 
pedpool
# Read in the IDs of the pedigrees sampled across the 2000 studies.
peds <- c(pvalres[,"studyped_1"], pvalres[,"studyped_2"],pvalres[,"studyped_3"])
# Check whether every pedigree in the pool of 55 was sampled
all(pedpool %in% peds) # every ped was sampled at least once
# See if pedigrees in the pool of 55 have been sampled roughly uniformly.
# There are 6000 study peds (2000 studies containing 3 peds each), so each
# pedigree should be sampled about 6000/55 = 109 times.
table(peds) # some sampled more than others by chance
sum(table(peds)) # Should be 6000=2000*3.
```

# Summary for paper: Type-1 error rates

* For the paper we require estimated type-1 error rates for each test.
* The global LR and global transmission tests depend on the
value of the carrier probability. Recall that we considered seven values
of the carrier probability,
leading to seven global LR and seven global transmission tests.
In addition there are the three local tests (local LR, RVS and
modified RVS) that don't depend on the carrier probability, for a total of 13 tests.

```{r}
true_carrier_prob <- 0.0032
carrier_probs <- true_carrier_prob*c(1/100,1/10,1/2,1,2,10,100)
tests <- c(paste0("globalLR",1:length(carrier_probs)),
           paste0("globaltrans",1:length(carrier_probs)),
           "localLR","RVS","modRVS")
tests
```
* In the code chunk below we loop over the tests and find
all the p-values for each one.
* For a given test, the simulation output from `simnull.R` 
includes p-values for each candidate cRV observed in each study. The different
cRVs are distinguished by the suffix `_1` for the first,
`_2` for the second and `_3` for the third.
    * Here "first", "second" and "third" refer to the order of the observed 
    candidate cRVs along chromosome 8, not to the size of the p-values.
    * When only two candidate cRVs are observed in a study the third p-values are `NA`
    and when there is only one candidate cRV observed in a study the second and third
    p-values are `NA`.
    * Note: Of the 2000 simulated studies, 1996 had only one observed candidate 
    cRV, four had two observed candidate cRVs and none of the simulated
    studies had three observed candidate cRVs. Thus, 2004
    candidate cRVs in total were tested. 

```{r}
# number of candidate cRVs in a study is the number of non-missing cRV IDs
cRV_IDs <- pvalres[,c("cRV_1","cRV_2","cRV_3")]
numcRVs <- apply(cRV_IDs,1,function(x) sum(!is.na(x)))
table(numcRVs)
```
    
* For a given test, the estimated type-1 error rate is the proportion
of the 2004 candidate-cRV tests that reject the null hypothesis.

```{r}
# Initialize matrix to hold all p-values. Rows of the matrix are
# studies and columns are tests.
pall <- matrix(NA,nrow=nrow(pvalres)*3,ncol=length(tests))
for(i in 1:length(tests)) {
  pvalcols <- paste0(tests[i],"_",1:3)
  p <- pvalres[,pvalcols] # p is a data frame of p-values for this test
  pall[,i] <- unlist(p) # vector of all p-values
}
colnames(pall) <- tests
type1errres<- function(pmat){
  ests <- apply(pmat,2,FUN=function(x) mean(x<=0.05,na.rm=TRUE))
  ses <- apply(pmat,2,FUN=se)
  res <- cbind(ests,ses); colnames(res) <- c("Estimates","SEs")
  return(res)
}
se <- function(x) { 
  n <- length(x)
  p <- mean(x<=0.05,na.rm=TRUE)
  return(sqrt(p*(1-p)/n))
}
```

## Estimated type-1 error rates

```{r}
round(type1errres(pall),3)
```

## Comments on results

* Type-1 error appears to be controlled at the 5\% level for all tests.
* However, all tests but the modified RVS appear to be conservative.
    * In particular, the LR tests (local LR, global LR and global transmission) 
and the RVS test have estimated type-1 error rates at 3 or more SEs  
below 0.05.
* The estimated type-1 error rates of the global tests (LR and transmission)
are similar across assumed values of the carrier probability within
10 times or less of the true carrier probability. 

## LaTeX table of results

* I organized the estimated type-1 error rates and their SEs into a 
LaTeX-formatted table similar to the table in Christina's thesis.
    * Note: The table is *not* automatically generated. It
    was done by hand from an empty template table (shown below) with the 
    cells of the table filled in by cutting-and-pasting from the R output.

\footnotesize
\begin{verbatim}
\begin{table}
\caption{Template table}
\begin{tabular}{lccccccc}
       & & \multicolumn{6}{c}{Method} \\ \cline{3-8}
       & & \multicolumn{3}{c}{Local} & & \multicolumn{2}{c}{Global} \\ \cline{3-5} \cline{7-8}
Factor$^*$ & & RVS & ModRVS & LR &   & LR & Transm. \\ \hline
NA & & e & e & e &    & -- & -- \\
1/100& & -- & -- & -- &  & e & e \\
1/10& & -- & -- & -- &  & e & e \\
1/2& & -- & -- & -- &  & e & e \\
1& & -- & -- & -- &  & e & e \\
2& & -- & -- & -- &  & e & e \\
10& & -- & -- & -- &  & e & e \\
100& & -- & -- & -- &  & e & e \\ \hline
\multicolumn{8}{l}{$^*$ Assumed carrier probability is factor times true value $p_c = 0.000320$.}
\end{tabular}
\end{table}
\end{verbatim}
\normalsize

\begin{table}[ht]
\centering
\caption{Estimated type-1 error rate (SE), all tested RVs (2004 in total)}
\begin{tabular}{lccccccc}
       & & \multicolumn{6}{c}{Method} \\ \cline{3-8}
       & & \multicolumn{3}{c}{Local} & & \multicolumn{2}{c}{Global} \\ \cline{3-5} \cline{7-8}
Factor$^*$ & & RVS & ModRVS & LR &   & LR & Transm. \\ \hline
NA & & 0.041 (0.003) & 0.052 (0.003) & 0.028 (0.002) &    & -- & -- \\
1/100& & -- & -- & -- &  & 0.037 (0.002) & 0.035 (0.002) \\
1/10& & -- & -- & -- &  & 0.037 (0.002) & 0.035 (0.002) \\
1/2& & -- & -- & -- &  & 0.036 (0.002) & 0.035 (0.002) \\
1& & -- & -- & -- &  & 0.036 (0.002) & 0.034 (0.002) \\
2& & -- & -- & -- &  & 0.036 (0.002) & 0.034 (0.002) \\
10& & -- & -- & -- &  & 0.033 (0.002) & 0.033 (0.002) \\
100& & -- & -- & -- &  & 0.007 (0.001) & 0.024 (0.002) \\ \hline
\multicolumn{8}{l}{$^*$ Assumed carrier probability is factor times true value $p_c = 0.000320$.}
\end{tabular}
\end{table}

\appendix
\renewcommand{\thesection}{A}

# Appendix: Format of the simulation-output file

* In the output files of p-values we include the following 
columns of information about the study replicate: the replicate number, 
the IDs of the three study pedigrees, and the IDs of the polymorphic
cRVs in each study.
* For $n_p$ carrier probabilities, we record $3\times(2\times n_p + 3)$
p-values: for each of the (up to) three cRVs tested we
have $n_p$ p-values for the global LR, $n_p$ for the global 
transmission approaches, and one each for the three local
approaches (local LR, RVS and modified RVS).
    * For the $i^{th}$ carrier probability and $j^{th}$ cRV, the 
    column names are `globalLR`$i$`_`$j$ for the global likelihood
    ratio test and `globaltrans`$i$`_`$j$ for the global transmission test.
    * For the local tests (which do not depend on the carrier probability),
    the column names for the $j^{th}$ cRVs are `localLR_`$j$, `RVS_`$j$ 
    and `modRVS_`$j$, respectively.
    * If there are fewer than three polymporhphic cRVs, the empty slots
    for the p-values are encoded as `NA`.
