---
title: "Simulate pool of pedigrees for simulation study"
output:
  pdf_document
date: "2023-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval=FALSE)
```

# Overview

* **NOTE** This .Rmd document is intended to be run on your PC and 
describes commands to simulate pedigrees in the
R script `simrvped.R`. 
* Simulating pedigrees is a compute-intensive task that should
be performed on the Compute Canada cluster. 
* Therefore, the first code
chunk in this RMarkdown document sets `eval=FALSE` for the entire 
document, to avoid having the R commands run when you knit on your PC. 
* At the end of the document there is a `purl()` command
that you must manually execute on your PC (i.e., cut-and-paste into the R console)
to generate the R script `simrvped.R`.
* After generating `simrvped.R` on your PC, port it to the Compute Canada 
cluster along with the SLURM script `simrvped.sh`. 
* Then submit the SLURM script to the cluster with the 
command `sbatch simrvped.sh`.

# Task-specific workflow

* We first load the `SimRVPedigree` package and a dataset of subtype-specific
hazard rates for Hodgkin's and non-Hodgkin's lymphoma.


```{r}
library(SimRVPedigree)
data(SubtypeHazards)
head(SubtypeHazards)

my_hazards <- hazard(SubtypeHazards,
                     subtype_ID = c("HL", "NHL"))
```

* We will run
an "array job" on the cluster, taking guidance from
Nirodha's SLURM scripts. 
* The R script `simRVped.R` will be called by the SLURM script 
a specified number of times (e.g. 150), and each call will
have its own "job ID" (e.g. 1:150).
* Each call of the R script can access its job ID through a 
Unix environment variable called `SLURM_ARRAY_TASK_ID`.
* The relevant code in the R script is:

```{r}
dID = Sys.getenv("SLURM_ARRAY_TASK_ID")
seed = as.numeric(dID)
# Set a seed value to assure the reproducibility.
if(!is.na(seed)) {
  set.seed(seed)
} else {
  warning("No task ID, setting seed to 1")
  set.seed(1)
}
```

* Nirodha wrote an R function `generatePeds()` to  call the `sim_RVped()` function
and write the output to a file whose name includes the job ID.
* The arguments to `sim_RVped()` are mostly as in Christina's thesis, Appendix C.6,
page 91,  with the exception of the carrier probability. For our simulations
of chromosome 8 the cumulative probability of the causal rare variants (cRVs) is 0.00016, and 
the `carrier_prob` should be twice this, or about 0.00032.
* The simulation program returns a full pedigree including all family members and also an ascertained version with only those family members recalled by the proband. We keep only the ascertained pedigree, writing it to a file in the  
`Outputfiles` directory. The ascertained-pedigree files 
for job ID `i` are called `full_pedi.txt`. 

```{r}
generatePeds = function(dataID){
  # Simulate pedigree ascertained for at least two individuals
  # affected by either Hodgkin's lymphoma or non-Hodgkin's lymphoma.
  out <- sim_RVped(hazard_rates = my_hazards,
                      GRR = c(35, 1),
                      RVfounder = TRUE,
                      FamID = 1,
                      founder_byears = c(1825, 1850),
                      ascertain_span = c(2000, 2010),
                      num_affected = 4,
                      stop_year = 2018,
                      carrier_prob = 0.00032, # 2x cum prob of cRVs
                      recall_probs = c(1, 1, 1, .75,.125,.125, 0),
                      first_diagnosis = 1980,
                      sub_criteria = list("HL",1)) # ascertain only if at least one HL

   write.table(out$ascertained_ped, file = paste0("Outputfiles/ascertained_ped",dataID,".txt"))
}
# Run the function.
generatePeds(dID)
```


## R script for the cluster.

* To create the R script `simrvped.R` for the Compute Canada cluster, cut-and-paste the following into the R command line on your PC:

```{r}
knitr::purl(input="simrvped.Rmd",output="simrvped.R")
```

* This command will return a file `simrvped.R` that includes every code chunk in this file `simrvped.Rmd`, including a code chunk at the very top of the file that sets `knitr` options.

* Delete the code chunk at the top of `simrvped.R` that sets the `knitr` options and the code chunk at the bottom that contains the `knitr::purl()` command.

* Then port `simrvped.Rmd` over to the Compute Canada cluster.
